Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-11-19 paper/graph.tex
Minor fixes to pagerank.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@666 3fc257bb-8550-0410-81b3-a5c103da83f7

The algorithm in Figure~\ref{fig:pr2} performs these iterations, using
linear-algebra concepts implemented in a MapReduce framework, with
data stored in 4 MapReduce objects: $A$, $MT$, $x$, and $y$.  $A$
stores the matrix, $MT$ its ``empty'' rows, while $x$ and $y$ are
vectors of length $|V|$.  The nonzeros of matrix $A$ are initialized
as described above; the PageRank vector $x$ is initialized uniformly.
A MapReduce object $MT$ contains the indices of all-zero rows of $A$,
corresponding to vertices with no outlinks.  The PageRank algorithm is
made more efficient by invoking the {\it aggregate()} operation once
on each of these MapReduce objects before the PageRank iterations
begin.  Since the 3 objects are all keyed by vertex, pre-aggregating
them moves all KV pairs with a given key to the same processor.  Many
of the subsequent PageRank operations then become local operations,
replacing a {\it collate()} (which is equivalent to an {\it
aggregate()} followed by {\it convert()}) with simply a {\it
convert()}.  In Figure~\ref{fig:pr2}, steps where {\it convert()} is
substituted for {\it collate()} are marked with an asterisk.  This
strategy of pre-aggregating to improve data locality is useful in many
MapReduce algorithms, but assumes that identical keys in different
MapReduce objects can be mapped to the same processor.

@@

Input: \\
\> Pre-aggregated MapReduce object $A$ = $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, A_{ij}]$ \\
\> Pre-aggregated MapReduce object $MT$ = indices of all-zero rows of $A$:  Key = $i$, Value = NULL \\
\> Pre-aggregated MapReduce object $x$ = initial PageRank vector:  Key = $i$, Value = $1/|V|$ \\
\> MapReduce object $y$ = vector, initially empty \\
\\
while $({residual} > {tolerance})$ \\
\> 1 Compute contribution for random jumps and zero-outdegree vertices \\
\> \> Add $x$ to $MT$ \\
$*$\> \> Convert $MT$:  Key is vertex $i$ \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$, MultiValue = $(x_i, NULL)$ if $i \in {MT}$ else MultiValue = $x_i$ \\
\> \> \> If ${nvalues} = 2$: $c_{dot} = c_{dot} + x_i$ \\
\> \> \> Output: Key = $i$, Value = NULL if ${nvalues} = 2$ \\
\> \> $c_{dot}$ = MPI\_Allreduce($c_{dot}$,MPI\_SUM) \\
\> 2 Compute $y = A^T x$  \\
\> \> Copy $x$ to $y$ \\
\> \> Add $A$ to $y$ \\
$*$\> \> Convert $y$:  Key is row index of $A$ \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$, Multivalue = $y_i, [j, A_{ij}]$ for nonzeros $A_{ij}$ in $A$ \\
\> \> \>  Output: Key = $j$, Value = $A_{ij} y_i$ \\
\> \> Collate $y$:  Key is column index of $A$ \\
\> \> Reduce $y$: \\
\> \> \> Input: Key = $j$, Multivalue = $A_{ij} y_i$ for nonzeroes $A_{ij}$ in column $j$ of $A$ \\
\> \> \> Output: Key = $j$, Value = $\sum_i A_{ij} y_i$ \\
\> 3 Add $c_{dot}$ to $y$; compute $y_{max}$ = max element in $y$ \\
\> \> Map $y$:  \\
\> \> \> Input: Key = $i$, Value = $y_i$ \\
\> \> \> $y_i = y_i + c_{dot}$; $y_{max} = {\rm max}(y_i,y_{max})$ \\
\> \> \> Output:  Key = $i$, Value = $y_i$ \\
\> \> $y_{max}$ = MPI\_Allreduce($y_{max}$,MPI\_MAX) \\
\> 4 Scale $y$ by $y_{max}$ \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$ \\
\> \> \> Output:  Key = $i$, Value = $y_i / y_{max}$ \\
\> 5 Compute residual \\
\> \> Add $y$ to $x$ \\
$*$\> \> Convert $x$:  Key is vertex $i$ \\
\> \> Reduce $x$:  \\
\> \> \> Input: Key = $i$, Multivalue = $x_i, y_i$ \\
\> \> \> $residual = {\rm max}(|x_i - y_i|,residual)$ \\
\> \> $residual$ = MPI\_Allreduce($residual$,MPI\_MAX) \\
\> 6 Copy $y$ to $x$ \\
\\
Output: \\
\> MapReduce object $x$ = PageRank vector: Key = $i$, Value = $x_i$ = numeric rank \\

@@

In Step 1 of a PageRank interation, $MT$ is effectively dotted with
$x$ to compute an adjustment $c_{dot}$ to the PageRank vector that
represents the uniform probability of going from a leaf page to any
other page.  $MT$ is unchanged by the operation.  Adjustments for
random jumps from any page are also computed.  In Step 2, the
matrix-vector product $A^T x$ is computed and the result stored in
$y$.  This is the most expensive part of the computation, requiring
two passes over the nonzero structure of $A$.  In the first pass, a
{\it convert()} gathers all local row entries $A_{ij}$ with their
associated $x_i$ entry, and a {\it reduce()} computes $A_{ij} x_i$.
In the second pass, a {\it collate()} gathers, for each $j$, all
contributions to the column sum $\sum_i A_{ij} x_i$, which is computed
by a second {\it reduce()}.  Steps 3 and 4 adjust and scale the
product vector; the work is proportional to $|V|$.  Step 5 computes
the residual as an infinity norm to determine whether the PageRank
vector has converged; work is again proportional to $|V|$.  Note that
$y$ and $x$ are both keyed by vertex at this point, so they can be
joined via an {\it add()} and {\it convert()} without need for the
communication required by a full {\it collate()}.  In Step 6, the
PageRank vector is overwritten prior to the next iteration.
Throughout the algorithm, we exploit the ability to call
MPI\_Allreduce to compute global dot products and norms.

This linear-algebra approach is attractive because it allows
construction of the MapReduce PageRank algorithm from simple kernels
(matrix-vector multiplication, dot products) on commonly used data
types (matrices, vectors).  One can build a variety of algorithms
based on these primitives; indeed, linear-algebra toolkits such as
Trilinos~\cite{Trilinos-Overview} and PETSc~\cite{petsc-efficient}
naturally exploit this idea.  In parallel, these toolkits share
pre-computed data layouts and communication maps among distributed
objects to enable efficient operations on the objects.  In a MapReduce
context, however, separating the data into vectors and matrices
results in additional overhead to gather KV pairs with the same key
into a key-multivalue (KMV) pair.  This overhead could be reduced in
the PageRank algorithm, for example, by storing the $x$ and $y$
vectors in a single MapReduce object, with key $i$ and value $(x_i,
y_i)$.  Doing this would eliminate the {\it copy()} in Step 2 of
Figure~\ref{fig:pr2}, as well as the {\it add()} and {\it convert()}
in Step 5.  A further performance improvement would store an entire
row of $A$ along with $x$ and $y$ in a single MapReduce object, with
key $i$ and value $(x_i, y_i, d_{out}, [j_1, A_{ij_1}], [j_2,
A_{ij_2}], \dots, [j_{d_{out}}, A_ij_{d_{out}}])$ (that is, storing
together, for each vertex, its old and new PageRank value along with
all out-going edges incident to the vertex).  This data layout would
eliminate an additional {it convert()}, but at the cost of code
reusability.  This data layout would be efficient for PageRank, but is
more complex and less likely to be useful for additional algorithms.
In addition, since each KV pair must fit within a single page of
memory (in MR-MPI), storing entire matrix rows in a single KV pair
limits the size of matrices to which the algorithm can be applied.
The more general approach used in Figure~\ref{fig:pr2} of storing
non-zeros separately, enables arbitrarily large problems to be solved.
The detailed timing break-downs in
section~\ref{subsec:results_pagerank} allow estimation of the benefit
of these possible enhancements.

>>>>

The algorithm in Figure~\ref{fig:pr2} performs these iterations, using
linear-algebra concepts implemented in a MapReduce framework, with
data stored in four MapReduce objects: $A$, $MT$, $x$, and $y$.  $A$
stores the matrix; $MT$ stores the row index of ``empty'' rows of $A$; 
and $x$ and $y$ are
vectors of length $|V|$.  The nonzeros of matrix $A$ are initialized
as described above; the PageRank vector $x$ is initialized uniformly.
A MapReduce object $MT$ contains the indices of all-zero rows of $A$,
corresponding to vertices with no outlinks.  The PageRank algorithm is
made more efficient by invoking the {\it aggregate()} operation once
on each of these MapReduce objects before the PageRank iterations
begin.  Since the objects are all keyed by vertex, pre-aggregating
them moves all KV pairs with a given key to the same processor.  Many
of the subsequent PageRank operations then become local operations,
replacing a {\it collate()} (which is equivalent to an {\it
aggregate()} followed by {\it convert()}) with simply a {\it
convert()}.  In Figure~\ref{fig:pr2}, steps where {\it convert()} is
substituted for {\it collate()} are marked with an asterisk.  This
strategy of pre-aggregating to improve data locality is useful in many
MapReduce algorithms, but assumes that identical keys in different
MapReduce objects can be mapped to the same processor.

@@

Input: \\
\> Pre-aggregated MapReduce object $A$ = $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, A_{ij}]$ \\
\> Pre-aggregated MapReduce object $MT$ = indices of all-zero rows of $A$:  Key = $i$, Value = NULL \\
\> Pre-aggregated MapReduce object $x$ = initial PageRank vector:  Key = $i$, Value = $1/|V|$ \\
\> MapReduce object $y$ = vector, initially empty \\
\> User-defined parameter $\alpha$ \\
\\
while $({residual} > {tolerance})$ \\
\> 1 Compute adjustment for random jumps and zero-outdegree vertices \\
\> \> Add $x$ to $MT$ \\
$*$\> \> Convert $MT$:  Key is vertex $i$ \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$, MultiValue = $(x_i, NULL)$ if $i \in {MT}$; else MultiValue = $x_i$ \\
\> \> \> If (${nvalues} = 2$) $c_{dot} = c_{dot} + x_i$; else $c_{dot} = c_{dot} + \alpha$ \\
\> \> \> Output: Key = $i$, Value = NULL if ${nvalues} = 2$ \\
\> \> $c_{dot}$ = MPI\_Allreduce($c_{dot}$, MPI\_SUM) / $|V|$ \\
\> 2 Compute $y = A^T x$  \\
\> \> Copy $x$ to $y$ \\
\> \> Add $A$ to $y$ \\
$*$\> \> Convert $y$:  Key is row index of $A$ \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$, Multivalue = $y_i, [j, A_{ij}]$ for nonzeros $A_{ij}$ in $A$ \\
\> \> \>  Output: Key = $j$, Value = $A_{ij} y_i$ \\
\> \> Collate $y$:  Key is column index of $A$ \\
\> \> Reduce $y$: \\
\> \> \> Input: Key = $j$, Multivalue = $A_{ij} y_i$ for nonzeroes $A_{ij}$ in column $j$ of $A$ \\
\> \> \> Output: Key = $j$, Value = $\sum_i A_{ij} y_i$ \\
\> 3 Add $c_{dot}$ to $y$; compute $y_{max}$ = max element in $y$ \\
\> \> Map $y$:  \\
\> \> \> Input: Key = $i$, Value = $y_i$ \\
\> \> \> $y_i = y_i + c_{dot}$; $y_{max} = {\rm max}(y_i,y_{max})$ \\
\> \> \> Output:  Key = $i$, Value = $y_i$ \\
\> \> $y_{max}$ = MPI\_Allreduce($y_{max}$, MPI\_MAX) \\
\> 4 Scale $y$ by $y_{max}$ \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$ \\
\> \> \> Output:  Key = $i$, Value = $y_i / y_{max}$ \\
\> 5 Compute residual \\
\> \> Add $y$ to $x$ \\
$*$\> \> Convert $x$:  Key is vertex $i$ \\
\> \> Reduce $x$:  \\
\> \> \> Input: Key = $i$, Multivalue = $x_i, y_i$ \\
\> \> \> $residual = {\rm max}(|x_i - y_i|,residual)$ \\
\> \> $residual$ = MPI\_Allreduce($residual$, MPI\_MAX) \\
\> 6 Copy $y$ to $x$ \\
\\
Output: \\
\> MapReduce object $x$ = PageRank vector: Key = $i$, Value = $x_i$ = numeric rank

@@

Step 1 of a PageRank interation is, effectively, a dot-product of $MT$ and
$x$ to compute an adjustment $c_{dot}$ to the PageRank vector that
represents random jumps and the uniform probability of going from a 
leaf page to any
other page.  $MT$ is unchanged by the operation.  
In Step 2, the
matrix-vector product $A^T x$ is computed and the result stored in
$y$.  This is the most expensive part of the computation, requiring
two passes over the nonzero structure of $A$.  In the first pass, a
{\it convert()} gathers all local row entries $A_{ij}$ with their
associated $x_i$ entry, and a {\it reduce()} computes $A_{ij} x_i$.
In the second pass, a {\it collate()} gathers, for each $j$, all
contributions to the column sum $\sum_i A_{ij} x_i$, which is computed
by a second {\it reduce()}.  Steps 3 and 4 adjust and scale the
product vector; the work is proportional to $|V|$.  Step 5 computes
the residual as an infinity norm to determine whether the PageRank
vector has converged; work is again proportional to $|V|$.  Note that
$y$ and $x$ are both keyed by vertex at this point, so they can be
joined via an {\it add()} and {\it convert()} without need for the
communication required by a full {\it collate()}.  In Step 6, the
PageRank vector is overwritten prior to the next iteration.
Throughout the algorithm, we exploit the ability to call
MPI\_Allreduce to compute global dot products and norms.

This linear-algebra approach is attractive because it allows
construction of the MapReduce PageRank algorithm from simple kernels
(matrix-vector multiplication, dot products) on commonly used data
types (matrices, vectors).  One can build a variety of algorithms
based on these primitives; indeed, linear-algebra toolkits such as
Trilinos~\cite{Trilinos-Overview} and PETSc~\cite{petsc-efficient}
naturally exploit this idea.  In parallel, each of these toolkits shares
pre-computed data layouts and communication maps among its distributed
objects to enable efficient operations on the objects.  In a MapReduce
context, however, separating the data into vectors and matrices
results in additional overhead to gather KV pairs with the same key
into a key-multivalue (KMV) pair.  This overhead could be reduced in
the PageRank algorithm, for example, by storing the $x$ and $y$
vectors in a single MapReduce object, with key $i$ and value $(x_i,
y_i)$.  Doing this would eliminate the {\it copy()} in Step 2 of
Figure~\ref{fig:pr2}, as well as the {\it add()} and {\it convert()}
in Step 5.  A further performance improvement would store an entire
row of $A$ along with $x$ and $y$ in a single MapReduce object, with
key $i$ and value $(x_i, y_i, d_{out}, [j_1, A_{ij_1}], [j_2,
A_{ij_2}], \dots, [j_{d_{out}}, A_ij_{d_{out}}])$ (that is, storing
together, for each vertex, its old and new PageRank value along with
all out-going edges incident to the vertex).  This data layout would
eliminate an additional {it convert()}, but at the cost of code
reusability.  This data layout would be efficient for PageRank, but is
more complex and less likely to be useful for additional algorithms.
In addition, since each KV pair must fit within a single page of
memory (in MR-MPI), storing entire matrix rows in a single KV pair
limits the size of matrices to which the algorithm can be applied.
The more general approach used in Figure~\ref{fig:pr2} of storing
non-zeros separately, enables arbitrarily large problems to be solved.
The detailed timing break-downs in
section~\ref{subsec:results_pagerank} allow estimation of the benefit
of these possible enhancements.
